---
title: "Concussions from 2012-2014 NFL Seasons"
author: "Serena Amro Gazze, Arjun Vadlamudi, Anirudhh Venkat"
date: "2025-02-17"
output: html_document
---

```{r libraries, echo = FALSE}

#Load libraries

suppressPackageStartupMessages(library(tidyverse))
library(ggplot2)
library(dplyr)
library(forcats)

```

```{r concussions, echo = FALSE}

#Load NFL Concussions dataset

concussions <- read.csv("/Users/anivenkat/Downloads/Concussion Injuries 2012-2014.csv")

concussions <- concussions %>%
  mutate(Position = ifelse(Position == "Comerback", "Cornerback", Position)) %>%
  filter(Position != "Long Snapper")

#Make new column for offensive vs defensive positions
offense_pos <- c("Center", "Offensive Tackle", "Quarterback", "Running Back", "Tight End", "Wide Receiver", "Guard", "Full Back")
defense_pos <- c("Defensive End", "Defensive Tackle", "Linebacker", "Safety", "Cornerback")

concussions <- concussions %>%
  mutate(Side = case_when(
    Position %in% offense_pos ~ "Offense",
    Position %in% defense_pos ~ "Defense"
  )) 

#head(concussions)
#str(concussions)

```

***Serena Amro Gazze: Analysis of NFL Concussions relative to Positions***

```{r positions, echo = FALSE}

#Filter Concussions dataset into Positions dataset
positions <- concussions %>% select(ID, Position, Side, Reported.Injury.Type, Unknown.Injury., Pre.Season.Injury., Weeks.Injured, Games.Missed, Play.Time.After.Injury)

#head(positions)

```

```{r unique, echo = FALSE}

#Write out all the unique positions
position <- positions %>% select(Position)

unique <- unique(position)
#unique

#str(position)

```

```{r offense, echo = FALSE}

#Make offense dataset
offense <- positions %>% filter(Side == "Offense")

#head(offense)
#str(offense)

```

```{r defense, echo = FALSE}

#Make defense dataset
defense <- positions %>% filter(Side == "Defense")

#head(defense)
#str(defense)

```

***Question 1a: Do offensive or defensive football positions result in more head injuries?***

```{r off vs def, echo = FALSE}

#Make data frame of numbers of offense vs defensive concussions
position_counts <- data.frame(
  Side = c("Offense", "Defense"),
  Count = c(nrow(offense), nrow(defense))
)

side_counts_plot <- ggplot(position_counts, aes(x = Side, y = Count, fill = Side)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5) +
  scale_fill_manual(values = c("Offense" = "blue", "Defense" = "red")) +
  labs(title = "Number of Concussions: Offense vs. Defense",
       x = "Position Group",
       y = "Count of Head Injuries") +
  theme_minimal()

side_counts_plot

```

***Question 1b: Which positions have the most and least head injuries?***

```{r positions plot, echo = FALSE}

#Make full positions plot
#Separate bars by color of offense vs defense

positions_plot <- ggplot(positions, aes(x = fct_infreq(Position), fill = Side)) +
  geom_bar(stat = "count") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("Offense" = "blue", "Defense" = "red")) + 
  labs(title = "Number of Concussions per Position",
       x = "Position",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

positions_plot

```

```{r most least injured, echo = FALSE}

#Make data frame of most injured positions
top_injured <- positions %>%
  count(Position, name = "Injury.Count") %>%
  arrange(desc(Injury.Count)) %>%
  slice_head(n = 5)

top_injured <- top_injured %>%
  mutate(Side = case_when(
    Position %in% offense_pos ~ "Offense",
    Position %in% defense_pos ~ "Defense",
  )) %>%
  select(Position, Side, Injury.Count)

#top_injured

#Make data frame of least injured positions
bottom_injured <- positions %>%
  count(Position, name = "Injury.Count") %>%
  arrange(Injury.Count) %>%
  slice_head(n = 5)

bottom_injured <- bottom_injured %>%
  mutate(Side = case_when(
    Position %in% offense_pos ~ "Offense",
    Position %in% defense_pos ~ "Defense",
  )) %>%
  select(Position, Side, Injury.Count)

#bottom_injured

```

***Question 1c: Which positions result in the top 5 numbers of head injuries?***

```{r most injured plot, echo = FALSE}

#Make top injured plot
top_plot <- ggplot(top_injured, aes(x = reorder(Position, -Injury.Count), y = Injury.Count, fill = Side)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Injury.Count), vjust = -0.5) +
  scale_fill_manual(values = c("Offense" = "blue", "Defense" = "red")) +
  labs(title = "Top 5 Positions with Most Head Injuries",
       x = "Position",
       y = "Count of Injuries") +
  theme_minimal()

top_plot

```

***Question 1d: Which positions result in the bottom 5 numbers of head injuries?***

```{r least injured plot, echo = FALSE}

#Make bottom injured plot
bottom_plot <- ggplot(bottom_injured, aes(x = reorder(Position, -Injury.Count), y = Injury.Count, fill = Side)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Injury.Count), vjust = -0.5) +
  scale_fill_manual(values = c("Offense" = "blue", "Defense" = "red")) +
  labs(title = "Top 5 Positions with Least Head Injuries",
       x = "Position",
       y = "Count of Injuries") +
  theme_minimal()

bottom_plot

```

***Question 1e: Which offensive positions result in the most and least head injuries?***

```{r offense plot, echo = FALSE}

#Make offensive positions plot
library(forcats)

offensive_plot <- ggplot(offense, aes(x = fct_infreq(Position), fill = Position)) +
  geom_bar(stat = "count") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Number of Concussions per Offensive Position",
       x = "Position",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

offensive_plot

```

***Question 1f: Which defensive positions result in the most and least head injuries?***

```{r defense plot, echo = FALSE}

#Make defensive positions plot
defensive_plot <- ggplot(defense, aes(x = fct_infreq(Position), fill = Position)) +
  geom_bar(stat = "count") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Number of Concussions per Defensive Position",
       x = "Position",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

defensive_plot

```

```{r injuries per season, echo = FALSE}

#Get number of injuries per season
twelve <- concussions %>% select(ID, Season, Position, Side, Reported.Injury.Type, Unknown.Injury., Pre.Season.Injury., Weeks.Injured, Games.Missed, Play.Time.After.Injury)
twelve <- twelve %>% filter(Season == "2012/2013")
twelve_count <- nrow(twelve)
#twelve_count

thirteen <- concussions %>% select(ID, Season, Position, Side, Reported.Injury.Type, Unknown.Injury., Pre.Season.Injury., Weeks.Injured, Games.Missed, Play.Time.After.Injury)
thirteen <- thirteen %>% filter(Season == "2013/2014")
thirteen_count <- nrow(thirteen)
#thirteen_count

fourteen <- concussions %>% select(ID, Season, Position, Side, Reported.Injury.Type, Unknown.Injury., Pre.Season.Injury., Weeks.Injured, Games.Missed, Play.Time.After.Injury)
fourteen <- fourteen %>% filter(Season == "2014/2015")
fourteen_count <- nrow(fourteen)
#fourteen_count

```

```{r season df, echo = FALSE}

#Make data frame of number of injuries per season
year_counts <- data.frame(
  Season = c("2012/2013", "2013/2014", "2014/2015"),
  Concussion_Count = c(twelve_count, thirteen_count, fourteen_count)
)

#year_counts

```

***Question 1g: Is there a trend in concussion frequency over different seasons (2012-2014 seasons)?***

```{r season plot, echo = FALSE}

#Plot concussions per season
year_plot <- ggplot(year_counts, aes(x = reorder(Season, -Concussion_Count), y = Concussion_Count, fill = Season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Concussion_Count), vjust = -0.5) + 
  labs(title = "Number of Concussions Per Season",
       x = "Season",
       y = "Concussion Count") +
  theme_minimal()

year_plot

```

```{r off vs def per season, echo = FALSE}

#Get number of specifically offense or defense positions per year
twelve_off <- twelve %>% filter(Side == "Offense")
twelve_off_count <- nrow(twelve_off)
#twelve_off_count

twelve_def <- twelve %>% filter(Side == "Defense")
twelve_def_count <- nrow(twelve_def)
#twelve_def_count

thirteen_off <- thirteen %>% filter(Side == "Offense")
thirteen_off_count <- nrow(thirteen_off)
#thirteen_off_count

thirteen_def <- thirteen %>% filter(Side == "Defense")
thirteen_def_count <- nrow(thirteen_def)
#thirteen_def_count

fourteen_off <- fourteen %>% filter(Side == "Offense")
fourteen_off_count <- nrow(fourteen_off)
#fourteen_off_count

fourteen_def <- fourteen %>% filter(Side == "Defense")
fourteen_def_count <- nrow(fourteen_def)
#fourteen_def_count

```

```{r off vs def df, echo = FALSE}

#Make data frames of number of specifically offense or defense positions per year
off_year_counts <- data.frame(
  Season = c("2012/2013", "2013/2014", "2014/2015"),
  Concussion_Count = c(twelve_off_count, thirteen_off_count, fourteen_off_count)
)

def_year_counts <- data.frame(
  Season = c("2012/2013", "2013/2014", "2014/2015"),
  Concussion_Count = c(twelve_def_count, thirteen_def_count, fourteen_def_count)
)

#off_year_counts
#def_year_counts

```

***Question 1h: Is there a trend in concussion frequency for offensive positions over different seasons (2012-2014)?***

```{r off per season plot, echo = FALSE}

#Make plot of numbers of offensive injuries per season
off_years_plot <- ggplot(off_year_counts, aes(x = Season, y = Concussion_Count, fill = Season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Concussion_Count), vjust = -0.5) + 
  labs(title = "Number of Offensive Concussions Per Season",
       x = "Season",
       y = "Concussion Count") +
  theme_minimal()

off_years_plot

```

***Question 1i: Is there a trend in concussion frequency for defensive positions over different seasons (2012-2014)?***

```{r def per season plot, echo = FALSE}

#Make plot of numbers of defensive injuries per season
def_years_plot <- ggplot(def_year_counts, aes(x = Season, y = Concussion_Count, fill = Season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Concussion_Count), vjust = -0.5) + 
  labs(title = "Number of Defensive Concussions Per Season",
       x = "Season",
       y = "Concussion Count") +
  theme_minimal()

def_years_plot

```

***Arjun Vadlamudi: Analysis of Injury Type relative to Seasons***
***Question 2a: How does the season correlate with injury type?***

```{r, echo = FALSE}

# Using group_by function to group data by season and injury type
injury_counts_by_season <- concussions %>%
  group_by(Season, Reported.Injury.Type) %>%
  count(Season, name = "InjuryCount")

# Create bar chart
ggplot(injury_counts_by_season, aes(x = Season, y = InjuryCount, fill = Reported.Injury.Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Injury Types per Season",
       x = "Season",
       y = "Number of Injuries",
       fill = "Injury Type") +
          theme_bw()

```

***Question 2b: How does injury type affect team's performance?***

```{r, echo = FALSE}

# Using group_by function to group data by Winning Team and injury type
injury_win_counts <- concussions %>%
  group_by(Winning.Team., Reported.Injury.Type) %>%
  count(Winning.Team., name = "InjuryCount")

# Create bar chart
ggplot(injury_win_counts, aes(x = Reported.Injury.Type, y = InjuryCount, fill = Winning.Team.)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Injury Type vs. Winning Team",
       x = "Injury Type",
       y = "Number of Injuries",
       fill = "Winning Team") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

***Anirudhh Venkat: Analysis of Teams Causing the Most Head Injuries***
***Question 3a: Which team has the most injuries?***

```{r, echo = FALSE}

# create "other" injuries
concussions <- concussions %>%
  mutate(Reported.Injury.Type = ifelse(Reported.Injury.Type == "",
                                       "Other", Reported.Injury.Type))

# get top 10 teams
top_teams <- concussions %>%
  group_by(Team) %>%
  summarize(TotalInjuries = n(), .groups = "drop") %>%
  arrange(desc(TotalInjuries)) %>%
  slice_head(n = 10)

# filter for top 10 teams and by injury
injury_counts_by_team_injury <- concussions %>%
  filter(Team %in% top_teams$Team) %>%
  group_by(Team, Reported.Injury.Type) %>%
  summarize(InjuryCount = n(), .groups = "drop") %>%
  left_join(top_teams, by = "Team")

# plot bar plot
ggplot(injury_counts_by_team_injury, aes(x = reorder(Team, -TotalInjuries), y = InjuryCount, fill = Reported.Injury.Type)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Teams with Most Injuries",
       x = "Team",
       y = "Number of Injuries",
       fill = "Injury Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

***Question 3b: Which team causes the most injuries?***

```{r, echo = FALSE}

# get top 10 injury counts by opposing teams
team_injury_count <- concussions %>%
  group_by(Opposing.Team) %>%
  summarise(Injury_Count = n()) %>%
  arrange(desc(Injury_Count)) %>%
  slice_head(n = 10)

# plot bubble chart
ggplot(team_injury_count, aes(x = reorder(Opposing.Team, -Injury_Count), y = Injury_Count, size = Injury_Count)) +
  geom_point(alpha = 0.7) +
  labs(title = "Injury Counts by Opposing Team",
       x = "Opposing Team",
       y = "Injury Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # control bubble sizes
  scale_size_continuous(range = c(5, 15))

```
